Inferring generation-interval distributions from contact tracing data: \\ \emph{in prep, PRSB}

### 2019-06-23 doc/network.tex
Some edits (went lightly only through Methods).

@@

\begin{flushleft}{
	\Large
	\textbf\newline{
		Inferring generation-interval distributions from contact tracing data
	}
}
\newline
\\
Sang Woo Park\textsuperscript{1,*}
David Champredon\textsuperscript{2,3}
Jonathan Dushoff\textsuperscript{2,4}
\\

@@

Generation intervals, defined as the time between when an individual is infected and when that individual infects another person, link two key quantities that describe an epidemic: the reproductive number, $\RR$, and the rate of exponential growth, $r$.
Often, generation intervals are measured through contact tracing by identifying who infected whom.
We study how observed intervals differ from ``intrinsic'' intervals that could be estimated by tracing individual-level infectiousness, and identify both spatial and temporal effects, including censoring (due to observation time), and the effects of susceptible depletion at various spatial scales. 
We show that correcting observed intervals for the effect of temporal censoring but \emph{not} for spatial effects provides a spatially informed ``effective'' generation-interval distribution, which correctly links $r$ and $\RR$.
TODO: Something about statistical methods?

@@

Assuming that the individual properties are independent of risk of infection, the population-level kernel is given by integrating over these individual variations across an ``aspect space'' $A$ which describes the population distributions:
\begin{equation}
K(\tau) = \int K_a (\tau) dA.
\end{equation}
\swp{David seems to prefer da over dA. I think I prefer to keep it as dA to represent the space of variation? Need to clarify this?}
\jd{dA is better, and we obviously need to clarify.}
\dc{So, isn't the standard way to write this integral like that: $ \int_{A} K_a (\tau) da$? Just trying to clarify here, this is confusing... }
The population-level kernel describes the rate at which secondary infections are expected to be caused by an an infected individual, on \emph{average}.

@@

For a single outbreak, the observed mean  generation interval through contact tracing will always be shorter than the mean intrinsic generation interval (Figure~\ref{fig:censor}).
There are two reasons for this phenomenon.
First, longer generation intervals are more likely to be missed due to right censoring (and short generation intervals are more likely to be observed).
In particular, when an epidemic is growing exponentially ($i(t) = i(0) \exp(rt)$), 
the censored generation-interval distribution (as well as the backward generation-interval distribution) can be written as the intrinsic generation-interval distribution discounted by the rate of exponential growth \citep{britton2019estimation}:
\begin{equation}
c_0(\tau) = b_0(\tau) \propto g(\tau) \exp(-r\tau),
\label{eq:exp}
\end{equation}
where $c_0$ and $b_0$ represent the ``initial'' censored and backward generation-interval distributions (during the exponential growth phase), respectively.
A deterministic simulation confirms that the censored generation-interval distribution has the same mean as the backward generation-interval distribution during this period (Figure~\ref{fig:censor}).
Second, decreasing number of susceptibles over the course of an epidemic makes long infections less likely to occur \citep{champredon2015intrinsic}.
As a result, even if contact tracing is performed through an entire epidemic, the mean intrinsic generation interval will be underestimated.
Long backward generation intervals near the end of an epidemic have little effect on the censored generation-interval distributions because they represent only a small portion of an outbreak.
Overall, we expect, naively using the observed generation-interval distribution, to underestimate the reproductive number.

@@

Throughout this paper, we assume that there is a constant per-pair contact rate $\lambda$ \citep{trapman2016inferring}.
In this case, the infectiousness of an individual $\RR_a = \int K_a(s) ds$ is the product of the number of acquaintances $N_a$ and the contact rate $\lambda$; the dilution term is equal to the reciprocal of the number of acquaintances: $\delta_a = 1/N_a$.
This assumption can be relaxed by allowing for asymmetry \citep{trapman2016inferring} or heterogeneity \citep{ball1997epidemics, ball2002general} in contact rates; 
for brevity, we do not pursue these directions here.

The population-level egocentric kernel is found by integrating over individual variations:
\begin{equation}\label{eq:ego}
\hat{K}(\tau) = \int \hat{K}_a(\tau) dA.
\end{equation}
\cite{trapman2016inferring} used this same kernel (also assuming a constant per-pair contact rate) to study the effect of network structure on the estimate of reproductive number.
The population-level egocentric generation-interval distribution is:
\begin{equation}
\hat{g}(\tau) = \frac{\hat{K}(\tau)}{\int \hat{K}(s) ds}.
\label{eq:conditional}
\end{equation}
The population-level egocentric generation-interval distribution describes the distribution of times at which secondary infections are realized by an \emph{average} primary case; for convenience, we will often omit ``population-level''.
Finally, the egocentric reproductive number can be defined by integrating over the egocentric kernel $\hat{K}(\tau)$ directly or by applying the Euler-Lotka equation to the egocentric generation-interval distribution \citep{trapman2016inferring}:
\dc{but that is only during the exponential growth phase, isn't it?}
\swp{Yes and no. Egocentric kernel is like an intrinsic kernel so there is no observation process attached to it.} 
\dc{mmh, still not super clear, but I guess it's fine for now; I don't really know how to improve this though... also the reviewers will be a good test for the clarity of this part...}
\begin{equation}
\frac{1}{\hat{\RR}} = \int \hat{g}(\tau) \exp(-r \tau) d\tau.
\label{eq:egorR}
\end{equation}
As the egocentric distribution always has a shorter mean than the intrinsic distribution, \Rhat\ will always be smaller than $\RR$ estimated from the intrinsic distribution;
this generation-interval-based argument provides a clear biological interpretation for the result presented by \cite{trapman2016inferring}.
% Similarly, \Rhat\ will be greater than the \emph{true} reproductive number \RR\, since it does not account for depletion of susceptibles by other routes.

@@

This calculation can be validated by simulating stochastic infection processes on a ``star'' network (i.e, a single infected individual at the center connected to multiple susceptible individuals who are not connected with each other).
Simulations confirm that in this case the distribution of \emph{contact} times matches the intrinsic generation-interval distribution, while the distribution of realized generation times (i.e., \emph{infection} times) matches the egocentric generation-interval distribution (\fref{local}, middle panel).

@@

In general, spatial reduction in the mean generation interval can be viewed as an effect of susceptible depletion and can be further classified into three levels: egocentric, local, and global.
Egocentric depletion, as discussed previously, is caused by an infected individual making multiple contacts to the same individual.
Local depletion refers to a depletion of susceptible individuals in a household or neighborhood;
we can think of these structures as small homogeneous networks embedded in a larger population structure (and therefore we can expect similar effects as we see in \fref{local}, right panel).
Both the egocentric and local depletion effects can be observed early in an epidemic, especially in a highly structured population, even if most of the population remains susceptible.
Finally, global depletion refers to overall depeletion of susceptibility as a population-level and explains the reduction in realized compared to intrinsic generation intervals that occurs even in a well-mixed population (\fref{censor}).

\section{Inferring generation-interval distributions from a contact tracing data}

@@

Since the right-censoring effect is a sampling bias, we want to take it into account explicitly.
In contrast, spatial effects have the same effect on how the epidemic spreads as they do on observed generation intervals. 
We therefore expect that if we correct observed generation intervals for the right-censoring effect, spatial components will be taken into account implicitly. 
When temporal correction is performed early in an outbreak, during the exponential growth phase, the effective distribution should incorporate egocentric and local spatial effects but not the global effects; we expect this distribution to correctly link $r$ and $\RR$.
Hereafter, we will refer to the temporally corrected distribution as the ``effective'' generation-interval distribution $\tsub{g}{eff}(\tau)$.
In a large homogeneously mixing population, the effective distribution will be equivalent to the intrinsic distribution.

\begin{figure}[!pbth]
\includegraphics[width=\textwidth]{../fig/cmp_reproductive.pdf}
\caption{\textbf{Comparison of estimates of reproductive number based on various methods.}
Using the observed generation-interval distributions (based on the first 1000 generation intervals) without correcting for right-censoring severely underestimates the reproductive number.
Similarly, using intrinsic and egocentric generation-interval distributions overestimates the reproductive number because they fail to account for local spatial effects.
Both population-level and individual-level methods provide estimates of reproductive number that are consistent with the empirical estimates, which we define as the average number of secondary cases generated by the first 100 infected individuals.
Boxplots are generated using 100 stochastic simulations of the SEIR model on an empirical network using Ebola-like parameters \citep{who2014ebola}: mean latent period $1/\sigma = 5 \textrm{ days}$ and mean infectious period $1/\gamma = 11.4 \textrm{ days}$. 
Per-pair contact rate $\lambda = 0.08 \textrm{ days}^{-1}$ is chosen to be sufficiently high such that differences are clear.
}
\label{fig:cmp}
\end{figure}

Here, we investigate two methods for correcting for temporal bias in a contact tracing data (see Methods for details).
We refer to the first method (Section~\ref{poplevel}) as the population-level method as it relies on the observed distribution aggregated across the entire population.
When an epidemic is growing exponentially, right-censoring causes the observed generation interval to be discounted by the exponential growth rate \eref{exp};
hence, we can ``undo'' the censoring by exponentially weighting the observed generation-interval distribution \citep{tomba2010some, nishiura2010time, britton2019estimation}:
\begin{equation}
\tsub{g}{eff}(\tau) \propto c_t(\tau) \exp(r\tau),
\label{eq:geff}
\end{equation}
where $r$ is the exponential growth rate.

@@

To test these methods, we simulate 100 epidemics with Ebola-like parameters on an empirical network \citep{leskovec2016snap}
and compare the estimates of reproductive number with empirical reproductive numbers, which we define as the average number of secondary cases generated by the first 100 infected individuals (\fref{cmp}).
As we predicted earlier, calculating reproductive number based on the intrinsic generation-interval distribution overestimates the empirical reproductive number;
estimates based on the egocentric generation-interval distribution \eref{conditional} are closer to empirical reproductive numbers but still suffer from overestimation as they do not account for indirect spatial effects. Direct estimates based on the observed generation intervals from contact tracing severely underestimates the empirical estimates.
While both population- and individual-level corrections provide similar estimates to the empirical reproductive number on average,
population-level estimates are more variable as they are more sensitive to outliers in generation intervals and our estimate of the growth rate.
For smaller values of \RR, we expect the differences to become smaller.
In Appendix, we present the same figure using smaller \RR\ (Appendix A.1) and using Erlang-distributed latent periods (Appendix A.2), which better corresponds to Ebola.
Overall, our qualitative conclusions do not change.

@@

The intrinsic generation-interval distribution, which describes the expected time distribution of secondary cases, provides a direct link between speed (exponential growth rate, $r$) and strength (reproductive number, $\RR$) of an epidemic \citep{wallinga2007generation, svensson2007note, svensson2015influence, park2019practical}.
However, observed generation-interval distributions can vary depending on how and when they are measured \citep{nishiura2010time, tomba2010some, champredon2015intrinsic, britton2019estimation};
determining which distribution correctly links $r$ and $\RR$ can be challenging.
Here, we analyze how observed generation intervals, measured through contact tracing, differ from intrinsic generation intervals.
Changes due to temporal censoring reflect observation bias, whereas changes due to spatial or network structure reflect the dynamics of the outbreak.
Thus correcting the observed distribution for temporal, but not spatial, effects provides the correct link between $r$ and $\RR$.

@@

We classified susceptible depletion into three levels: egocentric, local, and global.
Egocentric susceptible depletion refers to the effect of an infected individual making multiple contacts to the same susceptible individual.
Accounting for the egocentric effect allowed us to reinterpret the results by \cite{trapman2016inferring} using arguments based on generation intervals.
Local susceptible depletion refers to the effect of other closely related infected individuals (e.g., in a household or neighborhood) making multiple contacts to the same susceptible individual.
Global susceptible depletion refers to the overall decrease in the susceptible pool in the population.

@@

The generation-interval-based approaches to estimating the reproductive number often assumes that an epidemic grows exponentially \citep{wearing2005appropriate, wallinga2007generation, roberts2007model, park2019practical}.
In practice, heterogeneity in population structure can lead to subexponential growth \citep{szendroi2004polynomial, chowell2015western, chowell2016growing, chowell2016characterizing, kiskowski2016modeling, viboud2016generalized};
we therefore expect our simulations on an empirical network to be better characterized by subexponential growth models \citep{viboud2016generalized}.
However, our simulations suggest that the initial exponential growth assumption still provides a viable approach for estimating the reproductive number.

\swp{New paragraph:}
Contact tracing provides an effective way of collecting epidemiological data and controlling an outbreak \citep{clarke1998contact, eames2003contact, donnelly2003epidemiological}.
In particular, using tracing information allows us to infer real-time estimates of the time-varying reproductive number \citep{cauchemez2006estimating, hens2012robust, jewell2012enhancing, soetens2018real};
generation-interval distributions, which can be either assumed or estimated, often play a central role in describing these tracing data.
Our study illustrates that generation intervals measured through contact tracing data contain information about the underlying contact structure, which can be implicitly reflected in the estimates of the reproductive number;
this perspective can be particularly useful for characterizing an epidemic because detailed information about the contact structure is often unavailable.

Generation-interval distribution is a key, and often under-appreciated, component of disease modeling and forecasting.
Different definitions, and different measurement approaches, produce different estimates of these distributions.
We have shown that estimates based on observed generation intervals (e.g., through contact tracing) differ in predictable ways from estimates based on underlying measures of infectiousness (e.g., from shedding studies).
These predictable differences can arise from temporal effects, egocentric spatial effects, local spatial (or network) effects and population-level effects.
We have shown that correcting for temporal effects allows us to estimate a spatially informed ``effective'' distribution, which accurately describes how disease spreads in a population. 
Future studies should carefully consider how measurement influences estimated generation-interval distributions, and how these distributions influence the spread of disease.

@@

\subsection{Population-level method for estimating generation-interval distributions from contact tracing data}
\label{poplevel}

@@

\subsection{Individual-level method for estimating generation-interval distributions from contact tracing data}
\label{indlevel}

>>>>

\newcommand{\jdnew}{\jd{NEW}}
\newcommand{\jddel}[1]{\jd{DELETE: #1}}

@@

\begin{flushleft}{
	\Large
	\textbf\newline{
		Inferring generation-interval distributions from contact-tracing data
	}
}
\newline
\\
Sang Woo Park\textsuperscript{1,*}
David Champredon\textsuperscript{2,3}
Jonathan Dushoff\textsuperscript{2,4}
\\

@@

Generation intervals, defined as the time between when an individual is infected and when that individual infects another person, link two key quantities that describe an epidemic: the reproductive number, $\RR$, and the rate of exponential growth, $r$.
Often, generation intervals are measured through contact tracing by identifying who infected whom.
We study how observed intervals differ from ``intrinsic'' intervals that could be estimated by tracing individual-level infectiousness, and identify both spatial and temporal effects, including censoring (due to observation time), and the effects of susceptible depletion at various spatial scales. 
We develop and test statistical methods for both spatial and temporal corrections of generation intervals,
and show that correcting observed intervals for the effect of temporal censoring but \emph{not} for spatial effects provides a spatially informed ``effective'' generation-interval distribution, which correctly links $r$ and $\RR$.

@@

Assuming that the individual properties are independent of risk of infection, the population-level kernel is given by integrating over these individual variations \jdnew:
\begin{equation}
K(\tau) = \int K_a (\tau) f(a) da,
\end{equation}
The population-level kernel describes the rate at which secondary infections are expected to be caused by an an infected individual, on \emph{average}.
where $f(a)$ represents a probability density over a (possibly multi-dimensional) aspect space.
\dc{So, isn't the standard way to write this integral like that: $ \int_{A} K_a (\tau) da$? Just trying to clarify here, this is confusing... }
\jd{I haven't thought about this for a long time. dA actually represents something pretty confusing (a chunk of the \emph{distribution} across the space). Probably easier to change than explain, but we do need a distribution now.}

@@

For a single outbreak, the observed mean  generation interval through contact tracing will always be shorter than the mean intrinsic generation interval (Figure~\ref{fig:censor}).
There are two reasons for this phenomenon.
First, longer generation intervals are more likely to be missed due to right censoring (and short generation intervals are more likely to be observed).
In particular, when an epidemic is growing exponentially ($i(t) = i(0) \exp(rt)$), 
the censored generation-interval distribution (as well as the backward generation-interval distribution) can be written as the intrinsic generation-interval distribution discounted by the rate of exponential growth \citep{britton2019estimation}:
\begin{equation}
c_0(\tau) = b_0(\tau) \propto g(\tau) \exp(-r\tau),
\label{eq:exp}
\end{equation}
where $c_0$ and $b_0$ represent the ``initial'' censored and backward generation-interval distributions (during the exponential growth phase), respectively.
A deterministic simulation confirms that the censored generation-interval distribution has the same mean as the backward generation-interval distribution during this period (Figure~\ref{fig:censor}).
Second, the decreasing number of susceptibles over the course of an epidemic makes long infections less likely to occur \citep{champredon2015intrinsic}.
As a result, even if contact tracing is performed through an entire epidemic, the mean intrinsic generation interval will be underestimated.
Long backward generation intervals near the end of an epidemic have little effect on the censored generation-interval distributions because they represent only a small portion of an outbreak.
Overall, we expect, naively using the observed generation-interval distribution, to underestimate the reproductive number.

@@

Throughout this paper, we assume that there is a constant per-pair contact rate $\lambda$. \jddel{, following \citep{trapman2016inferring}}
In this case, the infectiousness of an individual $\RR_a = \int K_a(s) ds$ is the product of the number of acquaintances $N_a$ and the contact rate $\lambda$; the dilution term is equal to the reciprocal of the number of acquaintances: $\delta_a = 1/N_a$.
This assumption can be relaxed by allowing for asymmetry \citep{trapman2016inferring} or heterogeneity \citep{ball1997epidemics, ball2002general} in contact rates; 
for brevity, we do not pursue these directions here.

The population-level egocentric kernel is found by integrating over individual variations:
\begin{equation}\label{eq:ego}
\hat{K}(\tau) = \int \hat{K}_a(\tau) dA.
\end{equation}
\cite{trapman2016inferring} used this same kernel (also assuming a constant per-pair contact rate) to study the effect of network structure on the estimate of reproductive number.
The population-level egocentric generation-interval distribution is:
\begin{equation}
\hat{g}(\tau) = \frac{\hat{K}(\tau)}{\int \hat{K}(s) ds}.
\label{eq:conditional}
\end{equation}
The population-level egocentric generation-interval distribution describes the distribution of times at which secondary infections are realized by an \emph{average} primary case; for convenience, we will often omit ``population-level''.
Finally, the egocentric reproductive number can be defined by integrating over the egocentric kernel $\hat{K}(\tau)$ directly or by applying the Euler-Lotka equation to the egocentric generation-interval distribution \citep{trapman2016inferring}:
\dc{but that is only during the exponential growth phase, isn't it?}
\swp{Yes and no. Egocentric kernel is like an intrinsic kernel so there is no observation process attached to it.} 
\dc{mmh, still not super clear, but I guess it's fine for now; I don't really know how to improve this though... also the reviewers will be a good test for the clarity of this part...}
\jd{Not quite sure what Daniel is going for here: this equation depends on having an $r$, which does depend on exponentiality; also it is not how the number is ``defined'' at all. Not even sure if it's how it's estimated, except on a star network. This should be rewritten and not wait for reviewers.}
\begin{equation}
\frac{1}{\hat{\RR}} = \int \hat{g}(\tau) \exp(-r \tau) d\tau.
\label{eq:egorR}
\end{equation}
As the egocentric distribution always has a shorter mean than the intrinsic distribution, \Rhat\ will always be smaller than $\RR$ estimated from the intrinsic distribution;
this generation-interval-based argument provides a clear biological interpretation for the result presented by \cite{trapman2016inferring}.
% Similarly, \Rhat\ will be greater than the \emph{true} reproductive number \RR\, since it does not account for depletion of susceptibles by other routes.

@@

This calculation can be validated by simulating stochastic infection processes on a ``star'' network (i.e, a single infected individual at the center connected to multiple susceptible individuals who are not connected with each other).
Simulations (\fref{local}) confirm that in this case the distribution of \emph{contact} times matches the intrinsic generation-interval distribution (left panel), while the distribution of realized generation times (i.e., \emph{infection} times) matches the egocentric generation-interval distribution (middle panel).

@@

In general, spatial reduction in the mean generation interval can be viewed as an effect of susceptible depletion and can be further classified into three levels: egocentric, local, and global.
Egocentric depletion, as discussed previously, is caused by an infected individual making multiple contacts to the same individual.
Local depletion refers to a depletion of susceptible individuals in a household or neighborhood;
we can think of these structures as small homogeneous networks embedded in a larger population structure (and therefore we can expect similar effects as we see in \fref{local}, right panel).
Both the egocentric and local depletion effects can be observed early in an epidemic, especially in a highly structured population, even if most of the population remains susceptible.
Finally, global depletion refers to overall depeletion of susceptibility at the population level, and explains the reduction in realized compared to intrinsic generation intervals that occurs even in a well-mixed population (\fref{censor}).

\section{Inferring generation-interval distributions from contact-tracing data}

@@

Since the right-censoring effect is a sampling bias, we typically want to correct for it. 
In contrast, spatial effects have the same effect on how the epidemic spreads as they do on observed generation intervals. 
We therefore expect that starting from observed generation intervals and correcting for the right-censoring effect, will allow us to estimate an ``effective'' generation interval that accurately reflects dynamics of spread \jdnew. 
When temporal correction is performed early in an outbreak, during the exponential growth phase, the effective distribution should incorporate egocentric and local spatial effects but not the global effects; we expect this distribution to correctly link $r$ and $\RR$. 
We will call the temporally corrected effective generation-interval distribution $\tsub{g}{eff}(\tau)$.
In a large homogeneously mixing population, the effective distribution will be equivalent to the intrinsic distribution.

\begin{figure}[!pbth]
\includegraphics[width=\textwidth]{../fig/cmp_reproductive.pdf}
\caption{\textbf{Comparison of estimates of reproductive number based on various methods.}
Using the observed generation-interval distributions (based on the first 1000 generation intervals) without correcting for right-censoring severely underestimates the reproductive number.
Similarly, using the intrinsic generation-interval distribution overestimates the reproductive number because it fails to account for local spatial effects; the egocentric distribution corrects for this only partially \jdnew.
Both population-level and individual-level methods provide estimates of reproductive number that are consistent with the empirical estimates, which we define as the average number of secondary cases generated by the first 100 infected individuals, but the individual-level method is more precise \jdnew.
Boxplots are generated using 100 stochastic simulations of the SEIR model on an empirical network using Ebola-like parameters \citep{who2014ebola}: mean latent period $1/\sigma = 5 \textrm{ days}$ and mean infectious period $1/\gamma = 11.4 \textrm{ days}$. 
Per-pair contact rate $\lambda = 0.08 \textrm{ days}^{-1}$ is chosen to be sufficiently high such that differences are clear.
}
\label{fig:cmp}
\end{figure}

Here, we investigate two methods for correcting for temporal bias in contact-tracing data (see Methods for details).
We refer to the first method (Section~\ref{poplevel}) as the population-level method as it relies on the observed distribution aggregated across the entire population.
When an epidemic is growing exponentially, right-censoring causes the observed generation interval to be discounted by the exponential growth rate \eref{exp};
hence, we can ``undo'' the censoring by exponentially weighting the observed generation-interval distribution \citep{tomba2010some, nishiura2010time, britton2019estimation}:
\begin{equation}
\tsub{g}{eff}(\tau) \propto c_t(\tau) \exp(r\tau),
\label{eq:geff}
\end{equation}
where $r$ is the exponential growth rate.

@@

To test these methods, we simulate 100 epidemics with Ebola-like parameters on an empirical network \citep{leskovec2016snap}
and compare the estimates of reproductive number with empirical reproductive numbers, which we define as the average number of secondary cases generated by the first 100 infected individuals (\fref{cmp}).
As expected, calculating reproductive number based on the intrinsic generation-interval distribution overestimates the empirical reproductive number;
estimates based on the egocentric generation-interval distribution \eref{conditional} address this problem only partially \jdnew, as they do not account for indirect spatial effects. 
Direct estimates based on the observed generation intervals from contact tracing severely underestimate the empirical estimates.
While both population- and individual-level corrections provide similar estimates to the empirical reproductive number on average,
population-level estimates are more variable as they are more sensitive to outliers in generation intervals and our estimate of the growth rate.
For smaller values of \RR, we expect the differences to become smaller.
In Appendix, we present the same figure using smaller \RR\ (Appendix A.1) and using Erlang-distributed latent periods (Appendix A.2), which better corresponds to Ebola.
Overall, our qualitative conclusions do not change.

@@

The intrinsic generation-interval distribution, which describes the expected time distribution of secondary cases, provides a direct link between speed (exponential growth rate, $r$) and strength (reproductive number, $\RR$) of an epidemic \citep{wallinga2007generation, svensson2007note, svensson2015influence, park2019practical}.
However, observed generation-interval distributions can vary depending on how and when they are measured \citep{nishiura2010time, tomba2010some, champredon2015intrinsic, britton2019estimation};
determining which distribution correctly links $r$ and $\RR$ can be challenging.
Here, we analyze how observed generation intervals, measured through contact tracing, differ from intrinsic generation intervals.
Changes due to temporal censoring reflect observation bias, whereas changes due to spatial or network structure reflect the dynamics of the outbreak.
Thus, correcting the observed distribution for temporal, but not spatial, effects provides the correct link between $r$ and $\RR$.

@@

We think of susceptible depletion as operating on three levels: egocentric, local, and global.
Egocentric susceptible depletion refers to the effect of an infected individual making multiple contacts to the same susceptible individual.
Accounting for the egocentric effect allowed us to link the results by \cite{trapman2016inferring} to established results \jdnew\ based on generation intervals.
Local susceptible depletion refers to the effect of other closely related infected individuals (e.g., in a household or neighborhood) making multiple contacts to the same susceptible individual.
Global susceptible depletion refers to the overall decrease in the susceptible pool in the population.

@@

Generation-interval-based approaches to estimating the reproductive number often assume that an epidemic grows exponentially \citep{wearing2005appropriate, wallinga2007generation, roberts2007model, park2019practical}.
In practice, heterogeneity in population structure can lead to subexponential growth \citep{szendroi2004polynomial, chowell2015western, chowell2016growing, chowell2016characterizing, kiskowski2016modeling, viboud2016generalized};
we therefore expect our simulations on an empirical network to be better characterized by subexponential growth models \citep{viboud2016generalized}.
However, our simulations suggest that the initial exponential growth assumption still provides a viable approach for estimating the reproductive number.

Contact tracing provides an effective way of collecting epidemiological data and controlling an outbreak \citep{clarke1998contact, eames2003contact, donnelly2003epidemiological}.
In particular, using tracing information allows us to infer real-time estimates of the time-varying reproductive number \citep{cauchemez2006estimating, hens2012robust, jewell2012enhancing, soetens2018real};
generation-interval distributions, which can be either assumed or estimated, often play a central role in analyzing tracing data.
Our study illustrates that generation intervals measured through contact-tracing data contain information about the underlying contact structure, which can be implicitly reflected in the estimates of the reproductive number;
this perspective can be particularly useful for characterizing an epidemic because detailed information about the contact structure is often unavailable.

The generation-interval distribution is a key, and often under-appreciated, component of disease modeling and forecasting.
Different definitions, and different measurement approaches, produce different estimates of these distributions.
We have shown that estimates based on observed generation intervals (e.g., through contact tracing) differ in predictable ways from estimates based on underlying measures of infectiousness (e.g., from shedding studies).
These predictable differences can arise from temporal effects, egocentric spatial effects, local spatial (or network) effects and population-level effects.
We have shown that correcting observed intervals for temporal effects allows us to estimate a spatially informed ``effective'' distribution, which accurately describes how disease spreads in a population. 
Future studies should carefully consider how measurement influences estimated generation-interval distributions, and how these distributions influence the spread of disease.

@@

\subsection{Population-level method for estimating generation-interval distributions from contact-tracing data}
\label{poplevel}

@@

\subsection{Individual-level method for estimating generation-interval distributions from contact-tracing data}
\label{indlevel}
